---
title: |
  Title: Tumour size and overall survival in unifocal glioblastoma: a prognostic modelling and resampling study
author:
  - Kavi Fatania:
      institute: [DofRL, LIMR]
  - Russell Frood:
      institute: [DofRL, LIMR]
  - Hitesh Mistry:
      institute: [UofM]
  - Susan C. Short:
      institute: [DofCO]
  - James O'Connor:
      institute: [DofCS, DofRC, DofRI]
  - Andrew F. Scarsbrook:
      institute: [DofRL, LIMR]
  - Stuart Currie:
      institute: [DofRL, LIMR]
institute:
  - DofRL: Department of Radiology, Leeds Teaching Hospitals NHS Trust, UK
  - LIMR: Leeds Institute of Medical Research, University of Leeds, Leeds, UK
  - UofM: University of Manchester, Manchester, UK
  - DofCO: Department of Clinical Oncology, Leeds Teaching Hospitals NHS Trust, UK
  - DofCS: Division of Cancer Sciences, The University of Manchester, Manchester, UK
  - DofRC: Department of Radiology, The Christie Hospital, Manchester, UK
  - DofRI: Division of Radiotherapy and Imaging, Institute of Cancer Research, London, UK
journal: ""
date: ""
toc: false
output:
 bookdown::word_document2:
    fig_caption: yes
    pandoc_args:
    - --csl=Extras/european-radiology.csl
    - --citation-abbreviations=Extras/abbreviations.json
    - --filter=pandoc-crossref
    - --lua-filter=Extras/scholarly-metadata.lua
    - --lua-filter=Extras/author-info-blocks.lua
    - --reference-doc=/Users/Kavi/Radiology/PhD/Thesis/VolumeModelling/VolModWriteup/Extras/Reference_Document.docx
 bookdown::pdf_document2:
      pandoc_args:
      - --csl=Extras/european-radiology.csl
      - --citation-abbreviations=Extras/abbreviations.json
      - --filter=pandoc-crossref
      - --lua-filter=Extras/scholarly-metadata.lua
      - --lua-filter=Extras/author-info-blocks.lua
bibliography: "/Users/Kavi/MendeleyBibtexFiles/PhD.bib"
keywords: 
---

<br>

[Corresponding author]{.underline} 

Dr Kavi Fatania

Department of Radiology

Leeds General Infirmary

Great George Street

LS1 3EX

Kavi.fatania@nhs.net

ORCID: https://orcid.org/0000-0003-2421-1083

\pagebreak

**[Abstract]{.underline}**

[Background]{.underline}

An array of prognostic models of glioblastoma incorporating tumour volume or diameter exist - often with inconsistent results. Our aim was to investigate the relationship between overall survival (OS) and tumour size using an institutional dataset of patients with glioblastoma (GBM).

[Methods]{.underline}

Retrospectively identified patients with IDH-wildtype unifocal WHO grade 4 GBM from 2014-2020 with pre-operative imaging were included. Tumours were segmented using deep-learning and manual correction and diameter was measured manually. Association between core or whole tumour volume (CTV or WTV respectively) or diameter with OS was evaluated using univariable Cox proportional hazard models. Log transformation and penalised splines were used to assess non-linearity. Resampling with 1,000,000 repetitions and varying sample size was used to create Cox models for tumour size - percentage of significant models was calculated per sample size.

[Results]{.underline}

279 patients were included. Median (interquartile range) CTV 28.1cm^3^ (12.6 – 50.3), WTV 103.3cm^3^ (45.6 – 160.1) and diameter 4.4cm (3.3-5.4). Univariable models of volume or diameter ± log transformation showed no association with OS and spline functions did not show a non-linear trend. Resampling led to higher percentages of significant models (p\<0.05) when CTV or WTV was log transformed, and with larger samples. For n=50, 5.1% of WTV vs 5.2% log(WTV), 5.0% CTV vs 7.0% log(CTV) were significant. At n=279, 5.6% WTV and 9.7% log(WTV), 8.6% CTV vs 8.9% log(CTV) had p\<0.05. Larger samples increased the percentage of significant models for diameter. 

[Conclusion]{.underline}

Association between tumour size and OS was not demonstrated in the institutional dataset. Resampling, however, suggests that there may be a prognostic effect for size using larger samples, and after log transformation for volume.

Word count: 270

\pagebreak

**[Introduction]{.underline}**

Glioblastoma (GBM) is the most common primary brain malignancy in adults and patients have a median overall survival (OS) of 12-15 months despite maximal oncological treatment (maximum safe surgical resection followed by adjuvant radiotherapy with concurrent temozolomide and further 6 cycles of adjuvant temozolomide -- Stupp protocol)[@Stupp2005;@Stupp2009].


There has been great interest in investigating which prognostic factors are important for patients with GBM, and many prognostic models commonly include imaging biomarkers[@Tewarie2021]. Magnetic resonance imaging (MRI) is routinely performed for patients with GBM throughout their treatment journey and therefore is a popular modality for exploring prognostic imaging biomarkers. The diameter of the tumour core is routinely evaluated in clinical practice and is amongst the most common of the imaging predictors of OS to be investigated[@Henker2017]. There are now several methods available for automated or semi-automated segmentation of the different habitats of GBM, and it has therefore become more feasible to integrate tumour volumes into prognostic models, including in larger institutional datasets[@Henker2017;@Rudie2019;@Forghani2020].


Prognostic modelling studies have been inconsistent regarding the prognostic effect of tumour diameter and volume[@Henker2017,@Peeken2018;@Lacroix2001;@Compter2021;@Senders2020]. Some of this inconsistency may be due to the sample sizes that are used to derive the prognostic models. Large cohort studies of GBM have demonstrated a weak prognostic effect of tumour diameter[@Senders2020], but smaller studies have often resulted in mixed  results[@Compter2021;@Soltani2021;@Park2017a]. Similar findings are observed for tumour volume[@Henker2017;@Li2016;@Molinaro2020]. As well as variation in sample size, some of the inconsistency may be due to variations in handling continuous variables during the modelling process, for example leading to dichotomisation[@Wangaryattawanich2015], assumptions of a linear relationship to outcome[@Henker2017] and use of univariable model significance to select predictors into multivariable models[@Park2017a]. All these choices are known to impact upon the modelling process and may impact on the ability to determine accurate prognostic effects of the candidate predictors[@Harrell2015;@Riley2019b].


In this study, therefore, our aim is to use an institutional cohort of patients diagnosed with GBM to investigate the prognostic effect of tumour volume and diameter, and how choice of sample size and consideration of non-linear transformations may impact on the likelihood
of finding a prognostic effect using a resampling study.

<br>

**[Materials and methods]{.underline}**

*[Patient selection]{.underline}*

Neuro-oncology multidisciplinary team (MDT) records at our institution between 2014-2020 were retrospectively reviewed, and all patients (16 years and over) with histologically proven GBM were included. The catchment area includes 3.9 -- 4.4 million adults, and over this period,
3046 new primary brain neoplasms were discussed, with approximately 20% diagnosed with GBM or malignant glioma. Inclusion criteria were: MRI performed prior to any surgery, unifocal tumour (as determined by consultant neuroradiologist with \>10 years experience), and all four of
the following MRI sequences acquired: T1-weighted (T1W), T2-weighted (T2W), fluid attenuated inversion recovery (FLAIR) and gadolinium-enhanced T1W (T1Gd) sequence. Exclusion criteria were:
absence of pre-operative MRI, significant degradation of imaging due to artefact, or tumours were multifocal at presentation, documented IDH mutation on immunohistochemistry or cytogenetic testing.

<br>

*[Data preparation]{.underline}*

A summary of the data preparation is shown in **Figure \@ref(fig:flowchart)**. Digital Imaging and Communications in Medicine (DICOM) image retrieval was approved by the local institutional approval (REC ref: 19/YH/0300, IRAS project ID: 255585). DICOM image preparation was performed in python 3[@VanRossum2009]. DICOM images were retrieved from the institutional picture archive and communication system (PACS) and pseudonymised, and the image acquisition parameters are summarised in **Supplementary Table 1**. Images were converted to Neuroimaging Informatics Technology Initiative (NIfTI) file format using the dicom2nifti (v2.3.4) package[@Icometrix].

```{r flowchart, fig.cap="Flowchart summarising the preparation of imaging data for statistical analysis", echo=FALSE,fig.topcaption=TRUE}
knitr::include_graphics("figures/Fig1.tiff")
```

*FeTS = Federated Tumor Segmentation software, GBM = glioblastoma, IDHwt = isocitrate dehydrogenase wild-type, PACS = picture archive and communication system, T1Gd = gadolinium-enhanced T1-weighted imaging*

<br>

*[Image pre-processing and tumour segmentation]{.underline}*

Semi-automated tumour segmentations were produced using the Federated Tumor Segmentation (FeTS) software, an open-source platform with both command line interface (CLI) and graphical user interfaces (GUI), available for processing and segmentation of MRIs for patients with glioblastoma[@Pati2022]. Detailed description of the software, including packages and libraries used in FeTS is available elsewhere[@Pati2022] and utilises the same pre-processing steps used in the Multimodal Brain Tumor Image Segmentation Benchmark (BRATS) challenge[@Menze2015], and the
open-source software Cancer Imaging Phenomics Toolkit (CaPTk)[@Davatzikos2018]. The key 
features will be outlined.

The T2W, T1Gd and FLAIR sequences were rigidly co-registered first to the T1W sequence, then to the SRI24 brain atlas[@Rohlfing2009] and also spatially resampled to 1 x 1 x 1mm voxel resolution using the Greedy registration framework[@Yushkevich2016]. Images were then skull-stripped[@Thakur2020] and tumour segmentation was performed with the 'nnU-net' deep-learning network and pretrained model weights[@Isensee2021]. Tumours were automatically segmented into three volumes of interest
(VOIs). The three VOIs were defined as: i) necrotic tumour -- fluid signal intensity showing very high T2W signal and reduced T1Gd signal compared to the same area on T1W images; ii) enhancing tumour -- increased signal on T1Gd compared to the same area on T1W images and also increased T1Gd signal
compared to normal white matter regions on T1Gd images; iii) peritumoural oedema -- high FLAIR and T2W signal of the entire tumour, minus the necrotic and enhancing regions and not including ventricles or extra-axial CSF spaces[@Menze2015]. Tumour masks were used to produce two tumour volumes per patient: 1) core tumour volume (CTV, cm^3^) -- combination of necrotic and enhancing components; 2) whole tumour volume (WTV, cm^3^) -- CTV combined with the peritumoural oedema (see **Figure \@ref(fig:gbms)**).

```{r gbms, fig.cap="Definition of tumour volumes used in the study", echo=FALSE, fig.topcaption=TRUE}
knitr::include_graphics("figures/segmentation.tiff")
```

*Selected MRI axial images: left – fluid-attenuated inversion recovery (FLAIR); middle – gadolinium-enhanced T1-weighted (T1Gd); right – T1Gd image with overlay of core tumour segmentation (in green) and peritumoural oedema (yellow). Core tumour volume (CTV) defined as the enhancing and necrotic component of the tumour (green), and whole tumour volume (WTV) defined as the combination of core and peritumour oedema segmentation (green + yellow).*

<br>

The tissue segmentations were checked manually and corrected using the FeTS GUI. All segmentations were checked by an imaging fellow (5 years radiology experience). Independently, 50 segmentations were also checked by a consultant neuroradiologist (\>10 years consultant neuroradiology experience) and the inter-rater concordance compared using the dice similarity coefficient (DSC). The mean (± standard deviation) DSC for core and oedema segmentations was 0.94 ± 0.05 and 0.97 ± 0.03 respectively, which are equivalent to values published in the BRATS segmentation dataset, in which multiple experts raters segment the same GBM images, and our segmentation concordance was therefore within the expected variation of inter-rater agreement[@Menze2015].

Tumour diameter was defined as the maximum axial or craniocaudal diameter of the enhancing tumour core and was measured using the T1Gd sequence within imaging viewing software (Impax Version 6.5.3.3009, Agfa Healthcare) using in-built callipers on a submillimetre scale (mm - converted to cm).

<br>

*[Statistical analysis]{.underline}*

All statistical analysis was performed in `r R.version.string`. Association between CTV, WTV or tumour diameter with overall survival (OS) was investigated using univariable Cox regression
modelling. Hazard ratios (HRs), concordance indices and p-values for each model were used to assess performance. Any non-linear relationships between OS and size (volume or diameter) was explored using both logarithmic transformation and penalised spline function, the latter implemented using the 'pspline' function within the 'survival' package[@Therneau2023]. Penalised spline functions were used to assess for any trends in the data that might not be seen with a linear fit, as penalised splines allow a smooth curve to be fit to data[@Perperoglou2019]. Overfitting to the data points is discouraged by the inclusion of a roughness penalty, and the implementation doesn't require any pre-specification of the number of internal boundaries or knots. Model fits were assessed by plotting each tumour size parameter against the log-hazard ratio.

To assess the impact that log-transformation and sample size could have on potentially detecting a prognostic effect of tumour size on OS, we conducted a resampling study. Using different sample sizes (50, 100, 150, 200, 250, 279), bootstrapped samples were generated from the original dataset with replacement. A Cox regression model for each tumour size variable, both with and without log-transformation, was created and the model's p-value extracted. Bootstrapping was carried out for 1,000,000 repetitions at each sample size and for each of the tumour size variables. The percentage of models that had a p-value \<0.05, \<0.01 and \<0.001 was calculated across the 1,000,000 repetitions per sample size.

<br>

**[Results]{.underline}**

*[Demographics of the study population]{.underline}*

After exclusion of ineligible patients, 279 patients with GBM were included, with 236 deaths before the censor date of 31/10/2020. Demographic information for the GBM patients is summarised in **Table \@ref(tab:demotable)**. Median OS was 12 months (95% CI 11 -- 14 months), median age was 62 years (IQR 55 -- 68 years), 39% (108/279) of patients were female, and 74% (207/279) patients had a surgical resection of their GBM. Median (IQR) CTV was 28.1cm^3^ (12.6 -- 50.3), WTV was 103.3cm^3^ (45.6 -- 160.1) and tumour diameter was 4.4cm (3.3-5.4). Histograms of tumour size (**Supplementary Figure 1**) show that distributions of CTV and WTV are slightly positively skewed and tumour diameter is normally distributed prior to any transformation.

```{r, include=FALSE, cache=FALSE, echo=FALSE}
knitr::read_chunk('/Users/Kavi/Radiology/PhD/MachineLearning/data/MLscripts/Demographics.R')
```

```{r, ImportDemo, include=FALSE}
```

```{r demotable, tab.cap='Summary of demographic and treatment information about patients (n=279)', echo=FALSE}
ft_demo
```

<br>

*[Univariable Cox models of tumour size]{.underline}*

**Table \@ref(tab:volmods)** summarises the univariable Cox regression models for CTV, WTV and diameter, both with and without log-transformation. The results of the models derived from the institutional GBM images show limited evidence for a prognostic relationship between tumour volume or diameter and OS. C-indices for all models were 0.5, and all hazard ratios (HRs) cross 1.

```{r, include=FALSE, cache=FALSE}
knitr::read_chunk('/Users/Kavi/Radiology/PhD/MachineLearning/data/MLscripts/VolMod.R')
```

```{r, ImportVolMod, include=FALSE, echo=FALSE}
```

```{r, VolModSummary, include=FALSE, echo=FALSE}
```

```{r volmods, echo=FALSE, tab.cap = 'Summary of univariable Cox regression models for tumour size parameters'}
ft_vol
```

<br>

In **Supplementary Figure 2a-c**, each tumour size parameter (with and without log-transformation) is plot against the log-hazard ratio. The fit of a linear function to the data is compared with the use of a penalised splines, and these suggest that there is limited evidence that the tumour size parameters have a prognostic relationship - the model closely following the reference line for linear and non-linear functions.

<br>

*[Resampling study]{.underline}*

The results of the resampling study are shown in **Table \@ref(tab:simresults)**. For all size parameters, higher percentages of models with p\<0.05 were seen as the sample size increased, and for tumour volume (CTV or WTV), also after log transformation. 

```{r, SetupVolSimResults, include=FALSE, echo=FALSE}
```

```{r, TableVolSimResults, include=FALSE, echo=FALSE}
```

```{r simresults, tab.cap='Results of resampling study', echo=FALSE}
ft_volsim
```

<br>

For n=50, 5.1% WTV vs 5.2% log(WTV) and 5.0% CTV vs 7.0% log(CTV) models were significant. At n=279, 5.6% WTV vs 9.7% log(WTV) and 8.6% CTV vs 8.9% log(CTV) models were significant. At n=50 5.4% diameter vs 6.0% log(diameter) models were significant, and at n=279, 6.4% diameter vs 6.1% log(diameter) models were significant. **Supplementary Tables 2a-b** also show the percentages of models with p\<0.01 and p\<0.001, and this shows the same overall trend for the tumour size parameters, but with successively lower percentages of models as the p-value threshold for statistical significance is lowered.

**Figure \@ref(fig:boxplotdiam) - \@ref(fig:boxplotcore)** show the distributions of the p-values extracted from models during the resampling study. For CTV and WTV, but not diameter, there is downwards trend as sample size increases, suggesting that the probability of seeing a prognostic effect is increased as sample size increases. Results from the resampling study indicate that the effect of decreasing the p-value distribution, and thereby increasing the probability of demonstrating a prognostic effect of size, by log-transformation is greater for WTV than CTV across the range of sample sizes.


```{r, Boxplotfunction, include=FALSE, echo=FALSE}
```

```{r boxplotdiam,fig.cap='Boxplots of p-values from resampling study for tumour diameter models',echo=FALSE, dpi=300,fig.height=7.5,fig.width=10,out.height=450,out.width=600,fig.topcaption=TRUE}
GrpBxplot(diam,log.diam,Samp_sizes,"Tumour Diameter")
```

```{r boxplotwtv,fig.cap='Boxplots of p-values from resampling study for whole tumour volume (WTV) models', echo=FALSE,dpi=300,fig.height=7.5,fig.width=10,out.height=450,out.width=600,fig.topcaption=TRUE}
GrpBxplot(wtv,log.wtv,Samp_sizes,"Whole tumour volume")
```

```{r boxplotcore,fig.cap='Boxplots of p-values from resampling study for core tumour volume (CTV) models', echo=FALSE,dpi=300,fig.height=7.5,fig.width=10,out.height=450,out.width=600,fig.topcaption=TRUE}
GrpBxplot(core,log.core,Samp_sizes,"Core tumour volume")
```

*Each plot (Figure \@ref(fig:boxplotdiam) - \@ref(fig:boxplotcore)) shows the distribution of p-values extracted from each model created across the 1,000,000 repetitions for each sample size. For each tumour size parameter, the p-value distributions for models before and after log-transformation are shown side by side (see figure legend). Dotted lines indicate the 0.05 significance level.* 

<br>

**[Discussion]{.underline}**

Despite univariable models not demonstrating a prognostic relationship between size and OS in our institutional data, the resampling study suggests that there could be a prognostic role, and that this is more likely to be detected in larger studies. For WTV, log-transformation could also increase the probability of detecting a statistically significant effect. However, WTV is rarely explored as a
candidate prognostic variable in patients with GBM[@Henker2017,@Henker2019,@Iliadis2012]. In 65 patients, Iliadis et al. found no significant association between the volume of pre-operative 'edema - T2 abnormality (including the tumor)' (equivalent to WTV) and OS using univariable Cox modelling[@Iliadis2012]. It is unclear if any data transformation was considered. Other groups have investigated peritumoural oedema alone, rather than WTV, and these studies have shown mixed results[@Wangaryattawanich2015,@Henker2019,@Fuster-Garcia2018,@Zhang2014,@Choi2019]. Fuster-Garcia et al., for example, found no prognostic effect for peritumoural oedema volume in 84 patients[@Fuster-Garcia2018], whereas Wangaryattawanich et al. showed a statistically significant effect for 'edema' when dichotomising volume using a threshold of 85,000mm^3^ in a cohort of 94 patients[@Wangaryattawanich2015].

Tumour diameter and CTV are more commonly investigated[@Tewarie2021,@Henker2017] and there are several studies in similarly sized institutional datasets that also did not show a prognostic effect for either CTV[@Compter2021,@Henker2019,@Kickingereder2018] or diameter[@Park2017a,@Lutterbach2003]. However, a small positive effect size has been shown in studies with larger datasets[@Senders2020,@Li2016]. These findings support the resampling results, which indicates that CTV and diameter could potentially have prognostic effects, detected with larger samples. Li et al. for example found that contrast enhancing tumour volume had a small but statistically significant effect in a cohort of 1226 GBM patients (HR 1.004 95% CI 1.002-1.006, p \<0.001)[@Li2016].

An important consideration in regression modelling is the inclusion of any non-linear transformation of variables[@Steyerberg2009], although this is not routinely documented in prognostic modelling studies in GBM[@Tewarie2021,@Henker2017]. The advantages include more flexible modelling of continuous variables that might not have a simple linear relationship to outcome, but this comes with the drawback of potentially overfitting a model to the development dataset. In our resampling study, logarithmic transformation led to a higher rate of significant models for CTV and WTV, which could explain why some prognostic models that assume linear relationships between volume and outcome return non-significant results. The histograms for CTV and WTV show a small positive skew, whereas diameter is normally distributed and this could explain why log-transformation did not have similar impact on diameter in the resampling study.

Bootstrapping (resampling with replacement) as a method for exploring model uncertainty has been described elsewhere in the statistics literature[@Harrell2015;@Riley2019b,@Steyerberg2009] and was used in our resampling study to demonstrate the variability in the prognostic effect of tumour size. By resampling a dataset multiple times, researchers can identify the uncertainty in multiple aspects of the model building process, such as feature selection, internal validation of model accuracy and model stability[@Harrell2015]. Our study suggests that when selecting one of these size variables in GBM prognostic models based on univariable model significance, there could be up to 5-10% uncertainty in whether they might be included in the multivariable model. This is one of the reasons that this approach of univariable screening of candidate predictors is generally not recommended for multivariable model building within the modelling literature[@Harrell2015;@Riley2019b].

There are several limitations of our study. The MRI acquisition parameters were heterogenous, in particular the slice thickness, and this could have impacted upon the accuracy of volume measurements. However, the spatial resampling of images to an isotropic 1mm^3^ voxel resolution may have reduced the impact of acquisition heterogeneity. Furthermore, our dataset represents a retrospective clinical dataset, which in our institution's routine practice is likley to include different imaging acquisitions due to patients being referred from other centres, with their own (varying) MRI protocols. A proportion of our patients had to be excluded due to lack of the necessary MRI sequences for the deep-learning segmentation algorithm. The efficiency of a semi-automated segmentation approach outweighed the potential limititation of a reduced sample size. We investigated only three size variables but there are many others described in the literature. Whilst this could be deemed a limitation of our approach, we did not aim to provide a comprehensive study of the prognostic role of all possible tumour size parameter in GBM, but to investigate some of the methodological issues affecting this question, and that could be applied to any of the other continuous measures of tumour size in GBM. 

<br>

**[Conclusion]{.underline}**

In univariable models derived from our institutional dataset of patients with GBM, there was limited evidence to support a linear or non-linear prognostic association between volume or diameter and patient outcome. The resampling study, however, suggests that there could be a prognostic effect for CTV, WTV and diameter, but that it is more likely to be demonstrated using large datasets and, in the case of WTV, after log-transformation.

<br>

Word count: 2,512

<br>

[Acknowledgements]{.underline}

The authors would like to acknowledge Cancer Research UK funding for the Leeds Radiotherapy Research Centre of Excellence (RadNet; C19942/A28832).

[Funding]{.underline}

KF is a 4ward North Clinical PhD fellow funded by Wellcome award 203914/Z/16/Z. Salary for AFS & SC is supported by the Leeds Hospitals Charity and Leeds RadNET. Salary for JOC is supported by Cancer Research UK Advanced Clinician Scientist Fellowship (C19221/A22746).

<br>

## References
