---
title: |
  Title: Tumour size and overall survival in unifocal glioblastoma: a prognostic modelling and resampling study
author:
  - Kavi Fatania:
      institute: [DofRL, LIMR]
  - Russell Frood:
      institute: [DofRL, LIMR]
  - Hitesh Mistry:
      institute: [UofM]
  - Susan C. Short:
      institute: [LIMR, DofO]
  - James O'Connor:
      institute: [DofCS, DofRC, DofRI]
  - Andrew F. Scarsbrook:
      institute: [DofRL, LIMR]
  - Stuart Currie:
      institute: [DofRL, LIMR]
institute:
  - DofRL: Department of Radiology, Leeds Teaching Hospitals NHS Trust, UK
  - LIMR: Leeds Institute of Medical Research, University of Leeds, Leeds, UK
  - UofM: University of Manchester, Manchester, UK
  - DofO: Department of Oncology, Leeds Teaching Hospitals NHS Trust, UK
  - DofCS: Division of Cancer Sciences, The University of Manchester, Manchester, UK
  - DofRC: Department of Radiology, The Christie Hospital, Manchester, UK
  - DofRI: Division of Radiotherapy and Imaging, Institute of Cancer Research, London, UK
journal: ""
date: ""
toc: false
output:
 bookdown::word_document2:
    fig_caption: yes
    pandoc_args:
    - --csl=Extras/european-radiology.csl
    - --citation-abbreviations=Extras/abbreviations.json
    - --filter=pandoc-crossref
    - --lua-filter=Extras/scholarly-metadata.lua
    - --lua-filter=Extras/author-info-blocks.lua
    - --reference-doc=/Users/Kavi/Radiology/PhD/Thesis/VolumeModelling/VolModWriteup/Extras/Reference_Document.docx
 bookdown::pdf_document2:
      pandoc_args:
      - --csl=Extras/european-radiology.csl
      - --citation-abbreviations=Extras/abbreviations.json
      - --filter=pandoc-crossref
      - --lua-filter=Extras/scholarly-metadata.lua
      - --lua-filter=Extras/author-info-blocks.lua
bibliography: "/Users/Kavi/MendeleyBibtexFiles/PhD.bib"
keywords: 
---

<br>

[Corresponding author]{.underline} 

Dr Kavi Fatania\
Department of Radiology\
Leeds General Infirmary\
Great George Street\
LS1 3EX\
Kavi.fatania@nhs.net\
ORCID: https://orcid.org/0000-0003-2421-1083

\pagebreak

**[Abstract]{.underline}**

[Background]{.underline}

Various glioblastoma (GBM) prognostic models incorporating tumour volume or diameter have been proposed - often with inconsistent performance. The aim was to investigate the relationship between overall survival (OS) and tumour size using large, well curated institutional dataset of patients with GBM.

[Methods]{.underline}

Patients with IDH-wildtype unifocal WHO grade 4 GBM and pre-operative imaging between 2014-2020 were retrospectively reviewed. Tumours were segmented using deep-learning with manual correction and manual diameter measurements. Association between core (enhancing + necrotic tumour), whole (core + peritumoural oedema) volume (CV/WV) and diameter with OS was evaluated using univariable Cox proportional hazard models. Log transformation and penalised splines were used to assess non-linearity. Resampling with 1,000,000 repetitions and varying sample size was used to create Cox models for tumour size and percentage of significant models was calculated per sample size.

[Results]{.underline}

279 patients were included. Median (interquartile range) CV 28.1cm^3^ (12.6 – 50.3), WV 103.3cm^3^ (45.6 – 160.1) and diameter 4.4cm (3.3-5.4). Univariable models of volume or diameter ± log transformation showed no association with OS and spline functions did not show a non-linear trend. Resampling led to higher percentages of significant models (p<0.05) when CV or WV was log transformed, and with larger samples. For n=50, 5.1% of WV vs 5.2% log(WV), 5.0% CV vs 7.0% log(CV) were significant. At n=279, 5.6% WV and 9.7% log(WV), 8.6% CV vs 8.9% log(CV) had p<0.05. Larger samples increased the percentage of significant models for diameter.

[Conclusion]{.underline}

Although an association between tumour size and OS was not demonstrated in this large, well curated institutional dataset, resampling suggests a prognostic effect may become apparent if very large datasets and specific data transformations are utilised.

Word count: 270

\pagebreak

**[Introduction]{.underline}**

Glioblastoma (GBM) is the most common primary brain malignancy in adults and patients have a median overall survival (OS) of 12-15 months despite intensive oncological treatment (maximum safe surgical resection followed by adjuvant radiotherapy with concurrent temozolomide and further 6 cycles of adjuvant temozolomide -- Stupp protocol)[@Stupp2005;@Stupp2009].


There has been significant interest in investigating which prognostic factors are important for patients with GBM, in order to better stratify patients into different risk groups and due to the increasing interest in individualised treatment strategies ('personalised medicine')[@Mowforth2023]. Many such prognostic models include imaging biomarkers[@Tewarie2021]. Magnetic resonance imaging (MRI) is routinely performed for patients with GBM throughout their treatment pathway and is therefore a popular modality for exploring prognostic imaging biomarkers. The diameter of the tumour core, which is commonly defined as the enhancing and necrotic portion of the tumour[@Currie2022], is routinely evaluated in clinical practice and is amongst the most common of the imaging predictors of OS to be investigated[@Henker2017]. There are now several methods available for automated or semi-automated segmentation of the different tumour regions of GBM that are apparent on imaging[@Menze2015], such as the enhancing and necrotic tumour and peritumoural 'oedema', and it has therefore become more feasible to integrate tumour volumes into prognostic models, including in larger institutional datasets[@Henker2017;@Rudie2019;@Forghani2020].


Modelling studies have been inconsistent regarding the prognostic effect of tumour diameter and volume[@Henker2017;@Peeken2018;@Lacroix2001;@Compter2021;@Senders2020]. Some of this inconsistency may be due to the sample sizes that are used to derive the prognostic models. Large cohort studies of GBM have demonstrated a weak prognostic effect of tumour diameter[@Senders2020], but smaller studies have often resulted in mixed results[@Compter2021;@Soltani2021;@Park2017a]. Similar findings are observed for tumour volume[@Henker2017;@Li2016;@Molinaro2020]. As well as variation in sample size, some of the heterogeneity may be due to variations in handling continuous variables during the modelling process, for example leading to dichotomisation[@Wangaryattawanich2015], assumptions of a linear relationship to outcome[@Henker2017] and use of univariable model significance to select predictors into multivariable models[@Park2017a]. All these choices are known to impact upon the modelling process and may impact on the ability to determine accurate prognostic effects of the candidate predictors[@Harrell2015;@Riley2019b].


In this study, therefore, the aim is to use a large, well curated institutional cohort of patients diagnosed with GBM to investigate the prognostic effect of tumour volume and diameter, and to interrogate how choice of sample size and consideration of non-linear transformations may impact on the likelihood of finding a prognostic effect using a resampling study.

<br>

**[Materials and methods]{.underline}**

*[Ethical approval]{.underline}*

This was a retrospective study and therefore informed patient consent was not feasible. Ethical approval and institutional data access was approved via local ethical review committee (REC ref: 19/YH/0300, IRAS project ID: 255585).

<br>

*[Patient selection]{.underline}*

All patients (16 years and over) with histologically proven GBM treated at a single tertiary referral centre between 2014-2020 were retrospectively identified from neuro-oncology multidisciplinary team (MDT) records. The catchment area includes 3.9 – 4.4 million adults, and over this period, 3046 new primary brain neoplasms were discussed, with approximately 20% diagnosed with GBM or malignant glioma. Inclusion criteria were: MRI performed prior to any surgery, unifocal tumour (as determined by consultant neuroradiologist with \>10 years experience), and all four of the following MRI sequences acquired: T1-weighted (T1W), T2-weighted (T2W), fluid attenuated inversion recovery (FLAIR) and gadolinium-enhanced T1W (T1Gd) sequence. Exclusion criteria were: absence of pre-operative MRI, significant degradation of imaging due to artefact, or tumours were multifocal at presentation, documented IDH mutation on immunohistochemistry or cytogenetic testing.

<br>

*[Data preparation]{.underline}*

A summary of the data preparation is shown in **Figure \@ref(fig:flowchart)**. DICOM image preparation was performed in python 3.9[@VanRossum2009]. DICOM images were retrieved from the institutional picture archive and communication system (PACS) and pseudonymised, and the image acquisition parameters are summarised in **Supplementary Table 1**. Images were converted to Neuroimaging Informatics Technology Initiative (NIfTI) file format using the dicom2nifti (v2.3.4) package[@Icometrix].

```{r flowchart, fig.cap="Flowchart summarising the preparation of imaging data for statistical analysis", echo=FALSE,fig.topcaption=TRUE}
knitr::include_graphics("figures/Fig1.tiff")
```

*CV = Core Volume, FeTS = Federated Tumor Segmentation software, GBM = glioblastoma, IDHwt = isocitrate dehydrogenase wild-type, NifTI = Neuroimaging Informatics Technology Initiative, PACS = picture archive and communication system, T1Gd = gadolinium-enhanced T1-weighted imaging, WV = Whole Volume*

<br>

*[Image pre-processing and tumour segmentation]{.underline}*

Semi-automated tumour segmentations were produced using the Federated Tumor Segmentation (FeTS) software, an open-source platform with both command line interface (CLI) and graphical user interfaces (GUI), available for processing and segmentation of MRIs for patients with GBM[@Pati2022]. A detailed description of the software, including packages and libraries used in FeTS is available elsewhere[@Pati2022] and utilises the same pre-processing steps used in the Multimodal Brain Tumor Image Segmentation Benchmark (BRATS) challenge[@Menze2015], and the
open-source software Cancer Imaging Phenomics Toolkit (CaPTk)[@Davatzikos2018]. The key 
features are outlined below.

The T2W, T1Gd and FLAIR sequences were rigidly co-registered first to the T1W sequence, then to the SRI24 brain atlas[@Rohlfing2009] and also spatially resampled to 1 x 1 x 1mm voxel resolution using the Greedy registration framework[@Yushkevich2016]. Images were then skull-stripped[@Thakur2020] and tumour segmentation was performed with the 'nnU-net' deep-learning network and pretrained model weights[@Isensee2021]. Tumours were automatically segmented into three volumes of interest
(VOIs). The three VOIs were defined as: i) necrotic tumour -- fluid signal intensity showing very high T2W signal and reduced T1Gd signal compared to the same area on T1W images; ii) enhancing tumour -- increased signal on T1Gd compared to the same area on T1W images and also increased T1Gd signal
compared to normal white matter regions on T1Gd images; iii) peritumoural oedema -- high FLAIR and T2W signal of the entire tumour, minus the necrotic and enhancing regions and not including ventricles or extra-axial CSF spaces[@Menze2015]. Tumour masks were used to produce two tumour volumes per patient: 1) core volume (CV, cm^3^) -- combination of necrotic and enhancing components; 2) whole volume (WV, cm^3^) -- CV combined with the peritumoural oedema (see **Figure \@ref(fig:gbms)**).

```{r gbms, fig.cap="Definition of tumour volumes used in the study", echo=FALSE, fig.topcaption=TRUE}
knitr::include_graphics("figures/segmentation.tiff")
```

*Selected MRI axial images: left – fluid-attenuated inversion recovery (FLAIR); middle – gadolinium-enhanced T1-weighted (T1Gd); right – T1Gd image with overlay of core tumour segmentation (in green) and peritumoural oedema (yellow). Core volume (CV) defined as the enhancing and necrotic component of the tumour (green), and whole volume (WV) defined as the combination of core and peritumour oedema segmentation (green + yellow).*

<br>

The tissue segmentations were checked manually and corrected using the FeTS GUI. All segmentations were checked by a board-certified neuroradiology fellow (5 years radiology experience). Independently, 50 segmentations were also checked by a consultant neuroradiologist (\>10 years consultant neuroradiology experience) and the inter-rater concordance compared using the dice similarity coefficient (DSC). 

Tumour diameter was defined as the maximum axial or cranio-caudal diameter of the enhancing tumour core and was measured using the T1Gd sequence within imaging viewing software (Impax Version 6.5.3.3009, Agfa Healthcare) using in-built callipers on a submillimetre scale (mm - converted to cm) by two radiology trainees and checked by a board-certified neuroradiology fellow.

<br>

*[Statistical analysis]{.underline}*

All statistical analysis was performed in `r R.version.string`. Association between CV, WV or tumour diameter with overall survival (OS) was investigated using univariable Cox regression
modelling. Hazard ratios (HRs), concordance indices and p-values for each model were used to assess performance. Any non-linear relationships between OS and size (volume or diameter) was explored using both logarithmic transformation and penalised spline function, the latter implemented using the 'pspline' function within the 'survival' package[@Therneau2023]. Penalised spline functions were used to assess for any trends in the data that might not be seen with a linear fit, as penalised splines allow a smooth curve to be fit to data[@Perperoglou2019]. Overfitting to the data points is discouraged by the inclusion of a roughness penalty, and the implementation doesn't require any pre-specification of the number of internal boundaries or knots. Model fits were assessed by plotting each tumour size parameter against the log-hazard ratio.

To assess the impact that log-transformation and sample size could have on potentially detecting a prognostic effect of tumour size on OS, we conducted a resampling study. Using different sample sizes (50, 100, 150, 200, 250, 279), bootstrapped samples were generated from the original dataset with replacement. A Cox regression model for each tumour size variable, both with and without log-transformation, was created and the model's p-value extracted. Bootstrapping was carried out for 1,000,000 repetitions at each sample size and for each of the tumour size variables. The percentage of models that had a p-value \<0.05, \<0.01 and \<0.001 was calculated across the 1,000,000 repetitions per sample size. Any differences in the distribution of p-values between the sample sizes and between log-transformed and non-transformed models was assessed with two-sided Kolmogorov-Smirnov (KS) tests.

<br>

**[Results]{.underline}**

*[Demographics of the study population]{.underline}*

After exclusion of ineligible patients, 279 patients with GBM were included, with 236 deaths before the censor date of 31/10/2020. Demographic information for the GBM patients is summarised in **Table \@ref(tab:demotable)**. Median OS was 12 months (95% CI 11 -- 14 months), median age was 62 years (IQR 55 -- 68 years), 39% (108/279) of patients were female, and 26% (72/279) patients had a surgical biopsy of their GBM (with all remaining patients undergoing resection of their tumour to varying extent). Median (IQR) CV was 28.1cm^3^ (12.6 -- 50.3), WV was 103.3cm^3^ (45.6 -- 160.1) and tumour diameter was 4.4cm (3.3-5.4). Histograms of tumour size (**Supplementary Figure 1**) showed that distributions of CV and WV were slightly positively skewed and tumour diameter was normally distributed prior to any transformation. These data confirm that our population is representative of patients diagnosed with GBM in other typical neurosciences centres[@Verduin2021].

```{r, include=FALSE, cache=FALSE, echo=FALSE}
knitr::read_chunk('/Users/Kavi/Radiology/PhD/MachineLearning/data/MLscripts/Demographics.R')
```

```{r, ImportDemo, include=FALSE}
```

```{r demotable, tab.cap='Summary of patient demographics and treatment (n=279)', echo=FALSE}
ft_demo
```

<br>

*[Segmentations and Univariable Cox models of tumour size]{.underline}*

The mean (± standard deviation) DSC for core and oedema segmentations was 0.94 ± 0.05 and 0.97 ± 0.03 respectively, which are equivalent to values published in the BRATS segmentation dataset, in which multiple experts raters segment the same GBM images, and our segmentation concordance was therefore within the expected variation of inter-rater agreement[@Menze2015].

**Table \@ref(tab:volmods)** summarises the univariable Cox regression models for CV, WV and diameter, both with and without log-transformation. The results of the models derived from the institutional GBM images show limited evidence for a prognostic relationship between tumour volume or diameter and OS. C-indices for all models were 0.5, and all hazard ratios (HRs) cross 1.

```{r, include=FALSE, cache=FALSE}
knitr::read_chunk('/Users/Kavi/Radiology/PhD/MachineLearning/data/MLscripts/VolMod.R')
```

```{r, ImportVolMod, include=FALSE, echo=FALSE}
```

```{r, VolModSummary, include=FALSE, echo=FALSE}
```

```{r volmods, echo=FALSE, tab.cap = 'Summary of univariable Cox proportional hazards models for each tumour size parameter (Whole volume, core volume and diameter) for predicting overall survival (n=279)'}
ft_vol
```

<br>

In **Supplementary Figure 2a-c**, each tumour size parameter (with and without log-transformation) is plotted against the log-hazard ratio. The fit of a linear function to the data is compared with the use of a penalised splines, and these suggest that there is limited evidence that the tumour size parameters have a prognostic relationship - the model closely following the reference line for linear and non-linear functions. These results show that within our large, well curated institutional dataset, there is limited evidence of either a linear or non-linear prognostic relationship between OS and size using univariable Cox proportional hazards models.

<br>

*[Resampling study]{.underline}*

The results of the resampling study are shown in **Table \@ref(tab:simresults)**. For all size parameters, higher percentages of models with p\<0.05 were seen as the sample size increased, and for tumour volume (CV or WV), also after log transformation. 

```{r, SetupVolSimResults, include=FALSE, echo=FALSE}
```

```{r, TableVolSimResults, include=FALSE, echo=FALSE}
```

```{r simresults, tab.cap='Results of resampling study', echo=FALSE}
ft_volsim
```

<br>

For WV, 5.14 vs 5.60% (n=50 vs n=279), for CV 5.07 vs 8.60% and for diameter 5.43 vs 6.39% models had p-values\<0.05 across all repetitions. The distributions of p-values (across all repetitions per sample size at n=50 vs n=279) differed significantly on two-sided KS testing (all KS p-values\<0.0001). Considering log-transformation, the increase in models with p-values\<0.05 for WV versus log(WV) at n=50 (5.14 WV vs 5.20% log(WV)) was not reflected in a significantly different distribution of p-values on KS testing (KS p=0.92). However, there were significant differences observed in the distribution of p-values before and after log-transformation for CV at n=50 (5.07 vs 7.07%), and both CV (8.60 vs 8.94%) and WV (5.60 vs 9.70%) at n=279 (all KS p-values\<0.0001). For diameter, log-transformation did not lead to consistently increased percentages of models with p-values\<0.05. At n=50, 5.43 vs 5.95% (diameter vs log(diameter)) and at n=279, 6.39 vs 6.13% (diameter vs log(diameter)) models across all repetitions had p-values\<0.05.  **Supplementary Tables 2a-b** also show the percentages of models with p\<0.01 and p\<0.001, and this shows the same overall trend for the tumour size parameters, but with successively lower percentages of models as the p-value threshold is lowered.

**Figures \@ref(fig:boxplotdiam) - \@ref(fig:boxplotcore)** show the distributions of the p-values extracted from models during the resampling study. For CV and WV, but not diameter, there is a downwards trend as sample size increases, suggesting that the probability of seeing a prognostic effect is increased as sample size increases. Results from the resampling study indicate that the effect of decreasing the p-value distribution, and thereby increasing the probability of demonstrating a prognostic effect of size, by log-transformation is greater for WV than CV across the range of sample sizes. These results show that increasing sample size for all size parameters and, in the case of WV, log transformation may increase the chances of showing a significant association with OS. 


```{r, Boxplotfunction, include=FALSE, echo=FALSE}
```

```{r boxplotdiam,fig.cap='Boxplots showing the distribution of p-values (y-axis) extracted from each univariable model for tumour diameter vs overall survival created across the 1,000,000 repetitions for each sample size (x-axis). Boxes outline the interquartile range of p-values from the resampling experiment, with median values indicated by the central, thick black line. Tails represent the range of the distribution. Models with and without log-transformation are shown side by side (see figure legend). The dotted horizontal line represents the p-value threshold for statistical significance (0.05).',echo=FALSE, dpi=300,fig.height=7.5,fig.width=10,out.height=450,out.width=600,fig.topcaption=TRUE}
GrpBxplot(diam,log.diam,Samp_sizes,"Tumour Diameter")
```

```{r boxplotwtv,fig.cap='Boxplots showing the distribution of p-values (y-axis) extracted from each univariable model for whole volume vs overall survival created across the 1,000,000 repetitions for each sample size (x-axis). Boxes outline the interquartile range of p-values from the resampling experiment, with median values indicated by the central, thick black line. Tails represent the range of the distribution. Models with and without log-transformation are shown side by side (see figure legend). The dotted horizontal line represents the p-value threshold for statistical significance (0.05).', echo=FALSE,dpi=300,fig.height=7.5,fig.width=10,out.height=450,out.width=600,fig.topcaption=TRUE}
GrpBxplot(wtv,log.wtv,Samp_sizes,"Whole tumour volume")
```

```{r boxplotcore,fig.cap='Boxplots showing the distribution of p-values (y-axis) extracted from each univariable model for core tumour volume vs overall survival created across the 1,000,000 repetitions for each sample size (x-axis). Boxes outline the interquartile range of p-values from the resampling experiment, with median values indicated by the central, thick black line. Tails represent the range of the distribution. Models with and without log-transformation are shown side by side (see figure legend). The dotted horizontal line represents the p-value threshold for statistical significance (0.05).', echo=FALSE,dpi=300,fig.height=7.5,fig.width=10,out.height=450,out.width=600,fig.topcaption=TRUE}
GrpBxplot(core,log.core,Samp_sizes,"Core tumour volume")
```

<br>

**[Discussion]{.underline}**

We set out to investigate the prognostic effect of tumour volume and diameter in our institutional cohort of patient with GBM, and specifically to show how choice of sample size and consideration of non-linear transformations may impact on the chances of detecting a prognostic effect. Univariable models did not provide any evidence of a linear or non-linear prognostic relationship between size and OS in our institutional data, however the resampling study suggests that there could be a prognostic role, and that this is more likely to be detected in larger studies. For WV, log-transformation could also increase the probability of detecting a statistically significant effect, potentially due to the positively skewed distribution and WV may encompass more of the infiltrated brain tissue. However, WV is rarely explored as a candidate prognostic variable in patients with GBM[@Henker2017;@Henker2019;@Iliadis2012]. In 65 patients, Iliadis et al. found no significant association between WV (described as 'edema - T2 abnormality (including the tumor)') and OS using univariable Cox modelling[@Iliadis2012]. It is unclear if any log or other transformation was considered. Other groups have investigated peritumoural oedema alone, rather than WV, and these studies have shown mixed results[@Wangaryattawanich2015;@Henker2019;@Fuster-Garcia2018;@Zhang2014;@Choi2019]. Fuster-Garcia et al. found no prognostic effect for peritumoural oedema volume in 84 patients[@Fuster-Garcia2018], whereas Wangaryattawanich et al. showed a statistically significant effect for peritumoural oedema when dichotomising volume using a threshold of 85,000mm^3^ in a cohort of 94 patients[@Wangaryattawanich2015].

Tumour diameter and CV are more commonly investigated[@Tewarie2021;@Henker2017] and there are several studies in similarly sized institutional datasets that also did not show a prognostic effect for either CV[@Compter2021;@Henker2019;@Kickingereder2018] or diameter[@Park2017a;@Lutterbach2003]. However, a small positive effect size has been shown in studies with larger datasets[@Senders2020;@Li2016]. These findings support the resampling results, which indicates that CV and diameter could potentially have prognostic effects, detected with larger samples. Li et al. for example found that contrast enhancing tumour volume had a small but statistically significant effect in a cohort of 1226 GBM patients (HR 1.004 95% CI 1.002-1.006, p \<0.001)[@Li2016]. Whilst it could be argued that such a small univariable model effect size is not clinically significant, the multivariable effects are uncertain and size may have important interactions with other predictors. Future studies may be able to leverage multi-institutional networks[@Pati2022] or online imaging repositories to further increase statistical power for studying prognostic biomarkers.

An important consideration in regression modelling is the inclusion of any non-linear transformation of variables[@Steyerberg2009], although this is not routinely documented in prognostic modelling studies in GBM[@Tewarie2021;@Henker2017]. The advantages include more flexible modelling of continuous variables that might not have a simple linear relationship to outcome, but this comes with the drawback of potentially overfitting a model to the development dataset. In our resampling study, logarithmic transformation led to a higher rate of significant models for CV and WV, which could explain why some prognostic models that assume linear relationships between volume and outcome return non-significant results. In the present study, the histograms for CV and WV show a small positive skew, whereas diameter is normally distributed. This might explain why log-transformation did not have similar impact on diameter in our resampling study.

Bootstrapping (resampling with replacement) as a method for exploring model uncertainty has been described elsewhere in the statistics literature[@Harrell2015;@Riley2019b;@Steyerberg2009] and was used in our resampling study to demonstrate the variability in the prognostic effect of tumour size. By resampling a dataset multiple times, researchers can identify the uncertainty in multiple aspects of the model building process, such as feature selection, internal validation of model accuracy and model stability[@Harrell2015]. Our study suggests that when selecting one of these size variables in GBM prognostic models based on univariable model significance, there could be up to 5-10% uncertainty in whether they might be included in the multivariable model. This is one of the reasons that this approach of univariable screening of candidate predictors is generally not recommended for multivariable model building within the modelling literature[@Harrell2015;@Riley2019b].

There are several limitations of our study. The MRI acquisition parameters were heterogenous, in particular the slice thickness, and this could have impacted upon the accuracy of volume measurements. However, the spatial resampling of images to an isotropic 1mm^3^ voxel resolution should have reduced the impact of acquisition heterogeneity. Furthermore, the dataset represents a retrospective real-world clinical dataset, which in our institution's routine practice is likley to include different imaging acquisitions due to patients being referred from other centres, with their own (varying) MRI protocols. A proportion of our patients had to be excluded due to lack of the necessary MRI sequences for the deep-learning segmentation algorithm. The efficiency of a semi-automated segmentation approach outweighed the potential limitation of a reduced sample size. We investigated only three size variables but there are many others described in the literature. Whilst this could be deemed a limitation of our approach, we did not aim to provide a comprehensive study of the prognostic role of all possible tumour size parameters in GBM, but to investigate some of the methodological issues affecting this question, and that could be applied to any of the other continuous measures of tumour size in GBM. 

<br>

**[Conclusion]{.underline}**

Although univariable models derived from our large, well curated institutional dataset of patients with GBM showed limited evidence to support a linear or non-linear prognostic association between size and patient outcome, the resampling study suggests that there might be a prognostic effect, but that it is only likely to be demonstrated using very large datasets and, in the case of WV, after log-transformation.

<br>

Word count: 2,734

<br>

[Acknowledgements]{.underline}

The authors would like to acknowledge Cancer Research UK funding for the Leeds Radiotherapy Research Centre of Excellence (RadNet; C19942/A28832).

[Funding]{.underline}

KF is a 4ward North Clinical PhD fellow funded by Wellcome award (203914/Z/16/Z). AFS & SC receive salary support from a Leeds Hospitals Charity (9R01/1403) and Cancer Research UK (C19942/A28832) grants. Salary for JOC is supported by Cancer Research UK Advanced Clinician Scientist Fellowship (C19221/A22746).

<br>

## References
