---
title: |
  Title: Tumour size and overall survival in glioblastoma: a prognostic modelling and simulation study
author:
  - Kavi Fatania:
      institute: [DofRL, LIMR]
  - Russell Frood:
      institute: [DofRL, LIMR]
  - Hitesh Mistry:
      institute: [UofM]
  - Susan C. Short:
      institute: [DofCO]
  - James O'Connor:
      institute: [DofCS, DofRC, DofRI]
  - Andrew F. Scarsbrook:
      institute: [DofRL, LIMR]
  - Stuart Currie:
      institute: [DofRL, LIMR]
institute:
  - DofRL: Department of Radiology, Leeds Teaching Hospitals NHS Trust, UK
  - LIMR: Leeds Institute of Medical Research, University of Leeds, Leeds, UK
  - UofM: University of Manchester, Manchester, UK
  - DofCO: Department of Clinical Oncology, Leeds Teaching Hospitals NHS Trust, UK
  - DofCS: Division of Cancer Sciences, The University of Manchester, Manchester, UK
  - DofRC: Department of Radiology, The Christie Hospital, Manchester, UK
  - DofRI: Division of Radiotherapy and Imaging, Institute of Cancer Research, London, UK
journal: ""
date: ""
toc: false
output:
 bookdown::word_document2:
    fig_caption: yes
    pandoc_args:
    - --csl=Extras/clinical-mass-spectrometry.csl
    - --citation-abbreviations=Extras/abbreviations.json
    - --filter=pandoc-crossref
    - --lua-filter=Extras/scholarly-metadata.lua
    - --lua-filter=Extras/author-info-blocks.lua
    - --reference-doc=/Users/Kavi/Radiology/PhD/Thesis/VolumeModelling/VolModWriteup/Extras/Reference_Document.docx
 bookdown::pdf_document2:
      pandoc_args:
      - --csl=Extras/clinical-mass-spectrometry.csl
      - --citation-abbreviations=Extras/abbreviations.json
      - --filter=pandoc-crossref
      - --lua-filter=Extras/scholarly-metadata.lua
      - --lua-filter=Extras/author-info-blocks.lua
bibliography: "/Users/Kavi/MendeleyBibtexFiles/PhD.bib"
keywords: "CRAN, R, RMarkdown, RStudio, YAML"
---

<br>

[Corresponding author]{.underline} 

Dr Kavi Fatania

Department of Radiology

Leeds General Infirmary

Great George Street

LS1 3EX

Kavi.fatania@nhs.net

ORCID: https://orcid.org/0000-0003-2421-1083

\pagebreak

**[Abstract]{.underline}**

[Background]{.underline}

There are an array of prognostic models of glioblastoma incorporating tumour volume or diameter, developed with heterogeneous methodology and with variable results. Our aim was to investigate the relationship between overall survival (OS) and tumour size using an institutional dataset derived from patients with glioblastoma.

[Methods]{.underline}

Patients diagnosed with IDH-wildtype unifocal WHO grade 4 glioblastoma from 2014-2020, with pre-operative T1-weighted (T1W), T2-weighted (T2W), fluid attenuated inversion recovery (FLAIR) and post-gadolinium T1W (T1Gd) imaging were included. Tumours were semi-automatically segmented using deep-learning initially and then manually corrected. Maximum diameter was measured was also measured using in-house diagnostic image viewing software. Association between core or whole tumour volume (CTV or WTV respectively) or diameter with OS was evaluated using univariable Cox proportional hazard models. Log transformation and penalised spline functions were used to assess non-linearity. A resampling analysis  was performed using bootstrapping with varying sample sizes (50,100,150,200,250,279). The percentage of models with p-value <0.05 across repetitions was calculated per sample size.

[Results]{.underline}

279 patients were included. Median (interquartile range) CTV 28.1cm3 (12.6 – 50.3), WTV 103.3cm3 (45.6 – 160.1) and diameter 4.4cm (3.3-5.4). Univariable models of volume or diameter ± log transformation showed no association with OS and spline functions did not show a non-linear trend. The resampling analysis led to higher percentages of significant models when CTV or WTV was log transformed, and with larger sample sizes. For n=50, 5.1% of WTV vs 5.2% log(WTV), 5.0% CTV vs 7.0% log(CTV) had p<0.05. At n=279, 5.6% WTV and 9.7% log(WTV), 8.6% CTV vs 8.9% log(CTV) had p<0.05. Only increasing sample size increased the percentage of models with p<0.05 for tumour diameter. 

[Conclusion]{.underline}

There was no evidence of association between volume or diameter and patient outcome in a single institutional dataset derived from patients with glioblastoma. A resampling analysis suggests that there could be a prognostic effect for tumour size, and for WTV more likely demonstrated after log transformation and using larger datasets.

Abstract word count: 326

\pagebreak

**[Introduction]{.underline}**

Glioblastoma (GBM) is the most common primary brain malignancy in adults and patients have a median overall survival (OS) of 12-15 months despite maximal oncological treatment (maximum safe surgical resection followed by adjuvant radiotherapy with concurrent temozolomide and further 6 cycles of adjuvant temozolomide -- Stupp protocol)[@Stupp2005;@Stupp2009].


There has been great interest in investigating which prognostic factors are important for patients with GBM, and many prognostic models commonly include imaging biomarkers[@Tewarie2021]. Magnetic resonance imaging (MRI) is routinely performed for patients with GBM throughout their treatment journey and therefore is a popular modality for exploring prognostic imaging biomarkers. The diameter of the tumour core is routinely evaluated in clinical practice and is amongst the most common of the imaging predictors of OS to be investigated[@Henker2017]. There are now several methods available for automated or semi-automated segmentation of the different habitats of GBM, and it has therefore become more feasible to integrate tumour volumes into prognostic models, including in larger institutional datasets[@Henker2017;@Rudie2019;@Forghani2020].


Prognostic modelling studies have been inconsistent regarding the prognostic effect of tumour diameter and volume[@Henker2017,@Peeken2018;@Lacroix2001;@Compter2021;@Senders2020]. Some of this inconsistency may be due to the sample sizes that are used to derive the prognostic models. Large cohort studies of GBM have demonstrated a weak prognostic effect of tumour diameter[@Senders2020], but smaller studies have often resulted in mixed  results[@Compter2021;@Soltani2021;@Park2017a]. Similar findings are observed for tumour volume[@Henker2017;@Li2016;@Molinaro2020]. As well as variation in sample size, some of the inconsistency may be due to variations in handling continuous variables during the modelling process, for example leading to dichotomisation[@Wangaryattawanich2015], assumptions of a linear relationship to outcome[@Henker2017] and use of univariable model significance to select predictors into multivariable models[@Park2017a]. All these choices are known to impact upon the modelling process and may impact on the ability to determine accurate prognostic effects of the candidate predictors[@Harrell2015;@Riley2019b].


In this study, therefore, our aim is to use an institutional cohort of patients diagnosed with GBM to investigate the prognostic effect of tumour volume and diameter, and how choice of sample size and consideration of non-linear transformations may impact on the likelihood
of finding a prognostic effect using a resampling study.

<br>

**[Materials and methods]{.underline}**

*[Patient selection]{.underline}*

Neuro-oncology multidisciplinary team (MDT) records at our institution between 2014-2020 were retrospectively reviewed, and all patients (16 years and over) with histologically proven GBM were included. The catchment area includes 3.9 -- 4.4 million adults, and over this period,
3046 new primary brain neoplasms were discussed, with approximately 20% diagnosed with GBM or malignant glioma. Inclusion criteria were: MRI performed prior to any surgery, unifocal tumour (as determined by consultant neuroradiologist with \>10 years experience), and all four of
the following MRI sequences acquired: T1-weighted (T1W), T2-weighted (T2W), fluid attenuated inversion recovery (FLAIR) and gadolinium-enhanced T1W (T1Gd) sequence. Exclusion criteria were:
absence of pre-operative MRI, significant degradation of imaging due to artefact, or tumours were multifocal at presentation, documented IDH mutation on immunohistochemistry or cytogenetic testing.

<br>

*[Data preparation]{.underline}*
```{r, include=F}
library(captioner)
figures <- captioner(prefix = "Figure ", auto_space = FALSE)

figures("Fig1", "Flowchart summarising the preparation of imaging data for statistical analysis", display=FALSE)
```
A summary of the data preparation is shown in **Figure \@ref(fig:flowchart)**. Digital Imaging and Communications in Medicine (DICOM) image retrieval was approved by the local institutional approval (REC ref: 19/YH/0300, IRAS project ID: 255585). DICOM image preparation was performed in python 3[@VanRossum2009]. DICOM images were retrieved from the institutional picture archive and communication system (PACS) and pseudonymised, and the image acquisition parameters are summarised in **Supplementary Table 1**. Images were converted to Neuroimaging Informatics Technology Initiative (NIfTI) file format using the dicom2nifti (v2.3.4) package[@Icometrix].

```{r flowchart, fig.cap="Flowchart summarising the preparation of imaging data for statistical analysis", echo=FALSE}
knitr::include_graphics("figures/Fig1.tiff")
```

FeTS = Federated Tumor Segmentation software, GBM = glioblastoma, IDHwt = isocitrate dehydrogenase wild-type, PACS = picture archive and communication system, T1Gd = gadolinium-enhanced T1-weighted imaging

<br>

*[Image pre-processing and tumour segmentation]{.underline}*

Semi-automated tumour segmentations were produced using the Federated Tumor Segmentation (FeTS) software, an open-source platform with both command line interface (CLI) and graphical user interfaces (GUI), available for processing and segmentation of MRIs for patients with glioblastoma[@Pati2022]. Detailed description of the software, including packages and libraries used in FeTS is available elsewhere[@Pati2022] and utilises the same pre-processing steps used in the Multimodal Brain Tumor Image Segmentation Benchmark (BRATS) challenge[@Menze2015], and the
open-source software Cancer Imaging Phenomics Toolkit (CaPTk)[@Davatzikos2018]. The key 
features will be outlined.

The T2W, T1Gd and FLAIR sequences were rigidly co-registered first to the T1W sequence, then to the SRI24 brain atlas[@Rohlfing2009] and also spatially resampled to 1 x 1 x 1mm voxel resolution using the Greedy registration framework[@Yushkevich2016]. Images were then skull-stripped[@Thakur2020] and tumour segmentation was performed with the 'nnU-net' deep-learning network and pretrained model weights[@Isensee2021]. Tumours were automatically segmented into three volumes of interest
(VOIs). The three VOIs were defined as: i) necrotic tumour -- fluid signal intensity showing very high T2W signal and reduced T1Gd signal compared to the same area on T1W images; ii) enhancing tumour -- increased signal on T1Gd compared to the same area on T1W images and also increased T1Gd signal
compared to normal white matter regions on T1Gd images; iii) peritumoural oedema -- high FLAIR and T2W signal of the entire tumour, minus the necrotic and enhancing regions and not including ventricles or extra-axial CSF spaces[@Menze2015]. Tumour masks were used to produce two tumour volumes per patient: 1) core tumour volume (CTV, cm^3^) -- combination of necrotic and enhancing components; 2) whole tumour volume (WTV, cm^3^) -- CTV combined with the peritumoural oedema (see **Figure \@ref(fig:gbms)**).

```{r gbms, fig.cap="Definition of tumour volumes used in the study", echo=FALSE}
knitr::include_graphics("figures/segmentation.tiff")
```

<br>

Selected MRI axial images: left – fluid-attenuated inversion recovery (FLAIR); middle – gadolinium-enhanced T1-weighted (T1Gd); right – T1Gd image with overlay of core tumour segmentation (in green) and peritumoural oedema (yellow). Core tumour volume (CTV) defined as the enhancing and necrotic component of the tumour (green), and whole tumour volume (WTV) defined as the combination of core and peritumour oedema segmentation (green + yellow).

The tissue segmentations were checked manually and corrected using the FeTS GUI. All segmentations were checked by an imaging fellow (5 years radiology experience). Independently, 50 segmentations were also checked by a consultant neuroradiologist (\>10 years consultant neuroradiology experience) and the inter-rater concordance compared using the dice similarity coefficient (DSC). The mean (± standard deviation) DSC for core and oedema segmentations was 0.94 ± 0.05 and 0.97 ± 0.03 respectively, which are equivalent to values published in the BRATS segmentation dataset, in which multiple experts raters segment the same GBM images, and our segmentation concordance was therefore within the expected variation of inter-rater agreement[@Menze2015].

Tumour diameter was defined as the maximum axial or craniocaudal diameter of the enhancing tumour core and was measured using the T1Gd sequence within imaging viewing software (Impax Version 6.5.3.3009, Agfa Healthcare) using in-built callipers on a submillimetre scale (mm - converted to cm).

## References
